\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles

### 2023-11-08 Manuscript/manuscript.tex
Sync overleaf and Github

%\setmainfont{Helvetica}
\title{\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}
\author[1]{Boyi Guo}
\author[2]{\rr{Louise A. Huuki-Myers}}
\author[3,4]{\rr{Melissa Grant-Peters}}
\author[2]{\rr{Leonardo Collado-Torres}}
\author[1,5,6,7]{Stephanie C. Hicks\thanks{Correspondence to \url{shicks19@jhu.edu}}}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA} 
\affil[2]{Lieber Institute for Brain Development, Johns Hopkins Medical Campus, Baltimore, MD, USA}
\affil[3]{Genetics and Genomic Medicine, Great Ormond Street Institute of Child Health, University College London, London, UK}
\affil[4]{Aligning Science Across Parkinson’s (ASAP) Collaborative Research Network, Chevy Chase, MD, USA}
\affil[5]{Department of Biomedical Engineering, Johns Hopkins School of Medicine, Baltimore, MD, USA}
\affil[6]{Center for Computational Biology, Johns Hopkins University, Baltimore, MD, USA}
\affil[7]{Malone Center for Engineering in Healthcare, Johns Hopkins University, MD, USA}

@@

\subsection*{Availability and implementation}
The open source R package \texttt{escheR} is freely available on \rr{Bioconductor (\href{https://bioconductor.org/packages/escheR}{bioconductor.org/packages/escheR}).}
% and Zenodo (\fixme{add link}).

@@

\begin{figure*}[!t]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/insitu.jpg}
\caption{\small \textbf{\texttt{escheR} enables multi-dimensional spatial visualizations following the Gestalt principles of design}. (\textbf{A-B}) The traditional visualization displays the colocalization plot of the expression of two genes \textit{EFNA5} and \textit{EPHA5} (\textbf{A}) and the spatial domains from the dorsolateral prefrontal cortex in postmorteum human brain \cite{huukimyers_2023} (\textbf{B}) side-by-side, creating challenges to cognitively connecting colocalization status to spatial domains. (\textbf{C}) The watercolor effect enables displaying spatial domains by color-coding only outlines of circles. (\textbf{D}) \texttt{escheR} enables the multi-dimensional \textit{in situ} visualization that simultaneously displays the cortex layers and the colocalization status, substantially improving interpretability.}
\label{fig:visual} 
\end{center}
\end{figure*}

@@

Here, we provide an open-source package called \texttt{escheR} (named after the graphic artist M.C. Escher) in the R programming language \cite{R}, leading to a simplified interface to navigate the implementation of the multi-dimensional visualization in 2D space. By adapting \texttt{ggplot2} standard, \texttt{escheR} can be seamlessly integrated with \rr{popular data objects, including \texttt{SingleCellExperiment} \cite{amezquita2020} and \texttt{SpatialExperiment} \cite{righelli2022}, and easily work with popular pipelines, such as} \texttt{Seurat} \cite{hao_2021}, \texttt{Giotto} \cite{dries_2021} to name a few, and allow further theme customization with ease.

@@

Adding a third dimension to 2D plots has been a long-standing challenge in visualization \cite{odonoghue_2010}. Our proposal addresses this fundamental challenge by introducing simple but effective design principles. \rr{These principles lead to visually easier-to-interpret graphics compared to 2D-color gradients and geometry annotations (\textbf{Figure~\ref{fig:embedding}}).} Unlike computer-based interactive visualizations, the proposed visualization is free from any platform and technology restriction, creating an accessible and economical solution. In addition, the proposed visualization is easily scalable and hence can be applied to all types of spatially resolved data.

@@

\rr{
While the proposed design principle set-up a solid foundation, the proposed package \texttt{escheR} can be further optimized for image-based spatially resolved data \cite{caicedo_2017}. For example, to address possible overplotting of densely arranged single cells in tight space, strategies such as binning or hatching plots \cite{patrick2023spatial} can be incorporated. In addition, the flexibility to represent data points as polygons will be provided to allow visualization of cell morphology.
% Combining with a binning strategy, similar to \texttt{schex} \cite{freytag_2020} to avoid possible overplotting \rr{or hatching plot proposed in \texttt{lisaClust} \cite{patrick2023spatial}}, the proposed visualization can be also applied to visualized image-based spatially resolved data , in addition to aforementioned spot-based spatially resolved data (\textbf{Figure~\ref{fig:visual}}).  \rr{\texttt{escheR} could also be applied with kernel gene-weighted density plots from \texttt{Nebulosa} \cite{alquicirahernandez_2021} and other ggplot2-based visualizations.}
% \rr{Independently, \texttt{lisaClust} \cite{patrick2023spatial} visualize cell types within spatial domains by generating hatching plots for immunofluorescence and mass cytometry imaging data.} 
}

@@

%% EPILOGUE ================================================================
\section*{Abbreviations}

@@

\begin{itemize}[nosep]
    \item \textbf{Boyi Guo}:  Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing, Visualization
    \item \textbf{Erik D. Nelson}: Conceptualization, Software, Writing, Visualization
    \item \textbf{Louise A. Huuki-Myers}: Conceptualization, Software
    \item \textbf{Melissa Grant-Peters}: Conceptualization
    \item \textbf{Leonardo Collado-Torres}: Conceptualization, Software
    \item \textbf{Stephanie C. Hicks}: Conceptualization, Resources, Writing - Review \& Editing, Visualization, Supervision, Project administration, Funding acquisition
\end{itemize}

@@

\subsection*{Conflict of Interest} 
None declared.

\clearpage 
% \printbibliography
\bibliography{refs}

\clearpage
\onecolumn

{\huge Supplementary Materials}

\hrule

\vspace*{0.5cm}

\begin{center}

{\Large \texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}

\vspace*{0.75cm}

{\large Boyi Guo, Louise A. Huuki-Myers, Melissa Grant-Peters, Leonardo Collado-Torres, Stephanie C.\ Hicks$^*$}

\vspace*{0.3cm}

{\small $^*$Correspondence to \url{shicks19@jhu.edu}}

\end{center}

\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}
\setcounter{figure}{0}
\setcounter{table}{0}
\setcounter{section}{0}
\setcounter{page}{1}
\makeatletter
\renewcommand{\thefigure}{S\@arabic\c@figure}
\renewcommand{\thetable}{S\@arabic\c@table}
\renewcommand{\thesection}{Supplemental Note S\@arabic\c@section}
\makeatother

\vspace*{1cm}

{\bf \large Contents}

\begin{enumerate}
    \item \textbf{Supplemental Figures~S1.}
\end{enumerate}

\clearpage

\noindent {\LARGE Supplemental Figures}

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/embedding.jpg}
\caption{\small \textbf{\texttt{escheR} \rr{improves} multi-dimensional embedding visualizations}. The gene expression of \textit{POMGNT1} among peripheral blood mononuclear cells \cite{PBMC} under the UMAP representation. (\textbf{A}) The \texttt{schex} R/Bioconductor package uses color-coded convex hulls to annotate data-driven cell types, creating confusion when interpreting hexagons in overlapping hulls. (\textbf{B}) \texttt{escheR} plots hexagon-specific membership to avoid substantial overlapping of cluster membership. \rr{Due to the binning strategy, cautions in interpretation are needed due to possible cluster intermixing within each hexagon.} }
\label{fig:embedding} 
\end{center}
\end{figure*}

>>>>

%\setmainfont{Helvetica}
\title{\texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}
\author[1]{Boyi Guo}
\author[2]{Louise A. Huuki-Myers}
\author[3,4]{Melissa Grant-Peters}
\author[2]{Leonardo Collado-Torres}
\author[1,5,6,7]{Stephanie C. Hicks\thanks{Correspondence to \url{shicks19@jhu.edu}}}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA} 
\affil[2]{Lieber Institute for Brain Development, Johns Hopkins Medical Campus, Baltimore, MD, USA}
\affil[3]{Genetics and Genomic Medicine, Great Ormond Street Institute of Child Health, University College London, London, UK}
\affil[4]{Aligning Science Across Parkinson’s (ASAP) Collaborative Research Network, Chevy Chase, MD, USA}
\affil[5]{Department of Biomedical Engineering, Johns Hopkins School of Medicine, Baltimore, MD, USA}
\affil[6]{Center for Computational Biology, Johns Hopkins University, Baltimore, MD, USA}
\affil[7]{Malone Center for Engineering in Healthcare, Johns Hopkins University, MD, USA}

@@

\subsection*{Availability and implementation}
The open source R package \texttt{escheR} is freely available on Bioconductor (\href{https://bioconductor.org/packages/escheR}{bioconductor.org/packages/escheR}).
% and Zenodo (\fixme{add link}).

@@

\begin{figure*}[!t]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/insitu.jpg}
\caption{\small \textbf{\texttt{escheR} enables multi-dimensional spatial visualizations following the Gestalt principles of design}. (\textbf{A-B}) The traditional visualization displays the colocalization plot of the expression of two genes \textit{EFNA5} and \textit{EPHA5} (\textbf{A}) and the spatial domains from the dorsolateral prefrontal cortex in postmorteum human brain \cite{huukimyers_2023} (\textbf{B}) side-by-side, creating challenges to cognitively connecting colocalization status to spatial domains. (\textbf{C}) The watercolor effect enables displaying spatial domains by color-coding only outlines of circles. (\textbf{D}) \texttt{escheR} enables the multi-dimensional \textit{in situ} visualization that simultaneously displays the cortex layers and the colocalization status, substantially improving interpretability.}
\label{fig:visual} 
\end{center}
\vspace{-0.3in}
\end{figure*}

@@

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{Manuscript/figure/embedding.jpg}
\caption{\small \textbf{\texttt{escheR} improves multi-dimensional embedding visualizations}. The gene expression of \textit{POMGNT1} among peripheral blood mononuclear cells \cite{PBMC} under the UMAP representation. (\textbf{A}) The \texttt{schex} R/Bioconductor package uses color-coded convex hulls to annotate data-driven cell types, creating confusion when interpreting hexagons in overlapping hulls. (\textbf{B}) \texttt{escheR} plots hexagon-specific membership to avoid substantial overlapping of cluster membership. Due to the binning strategy, cautions in interpretation are needed due to possible cluster intermixing within each hexagon.} 
\label{fig:embedding} 
\end{center}
\vspace{-0.3in}
\end{figure*}

Here, we provide an open-source package called \texttt{escheR} (named after the graphic artist M.C. Escher) in the R programming language \cite{R}, leading to a simplified interface to navigate the implementation of the multi-dimensional visualization in 2D space. By adapting \texttt{ggplot2} standard, \texttt{escheR} can be seamlessly integrated with popular data objects, including \texttt{SingleCellExperiment} \cite{amezquita2020} and \texttt{SpatialExperiment} \cite{righelli2022}, and easily work with popular pipelines, such as \texttt{Seurat} \cite{hao_2021}, \texttt{Giotto} \cite{dries_2021} to name a few, and allow further theme customization with ease.

@@

Adding a third dimension to 2D plots has been a long-standing challenge in visualization \cite{odonoghue_2010}. Our proposal addresses this fundamental challenge by introducing simple but effective design principles. These principles lead to visually easier-to-interpret graphics compared to 2D-color gradients and geometry annotations (\textbf{Figure~\ref{fig:embedding}}). Unlike computer-based interactive visualizations, the proposed visualization is free from any platform and technology restriction, creating an accessible and economical solution. In addition, the proposed visualization is easily scalable and hence can be applied to all types of spatially resolved data.

@@

While the proposed design principle set-up a solid foundation, the proposed package \texttt{escheR} can be further optimized for image-based spatially resolved data \cite{caicedo_2017}. For example, to address possible overplotting of densely arranged single cells in tight space, strategies such as binning or hatching plots \cite{patrick2023spatial} can be incorporated. In addition, the flexibility to represent data points as polygons will be provided to allow visualization of cell morphology.
% Combining with a binning strategy, similar to \texttt{schex} \cite{freytag_2020} to avoid possible overplotting \rr{or hatching plot proposed in \texttt{lisaClust} \cite{patrick2023spatial}}, the proposed visualization can be also applied to visualized image-based spatially resolved data , in addition to aforementioned spot-based spatially resolved data (\textbf{Figure~\ref{fig:visual}}).  \rr{\texttt{escheR} could also be applied with kernel gene-weighted density plots from \texttt{Nebulosa} \cite{alquicirahernandez_2021} and other ggplot2-based visualizations.}
% \rr{Independently, \texttt{lisaClust} \cite{patrick2023spatial} visualize cell types within spatial domains by generating hatching plots for immunofluorescence and mass cytometry imaging data.}

@@

% \printbibliography
\bibliography{refs}

\clearpage
%% EPILOGUE ================================================================
\section*{Abbreviations}

@@

\begin{itemize}[nosep]
    \item \textbf{Boyi Guo}:  Conceptualization, Methodology, Software, Validation, Formal analysis, Investigation, Data Curation, Writing, Visualization
    \item \textbf{Louise A. Huuki-Myers}: Conceptualization, Software
    \item \textbf{Melissa Grant-Peters}: Conceptualization
    \item \textbf{Leonardo Collado-Torres}: Conceptualization, Software
    \item \textbf{Stephanie C. Hicks}: Conceptualization, Resources, Writing - Review \& Editing, Visualization, Supervision, Project administration, Funding acquisition
\end{itemize}

@@

\subsection*{Conflict of Interest} 
None declared.
% \onecolumn

% {\huge Supplementary Materials}

% \hrule

% \vspace*{0.5cm}

% \begin{center}

% {\Large \texttt{escheR}: Unified multi-dimensional visualizations with Gestalt principles}

% \vspace*{0.75cm}

% {\large Boyi Guo, Louise A. Huuki-Myers, Melissa Grant-Peters, Leonardo Collado-Torres, Stephanie C.\ Hicks$^*$}

% \vspace*{0.3cm}

% {\small $^*$Correspondence to \url{shicks19@jhu.edu}}

% \end{center}

% \renewcommand{\figurename}{Supplementary Figure}
% \renewcommand{\tablename}{Supplementary Table}
% \setcounter{figure}{0}
% \setcounter{table}{0}
% \setcounter{section}{0}
% \setcounter{page}{1}
% \makeatletter
% \renewcommand{\thefigure}{S\@arabic\c@figure}
% \renewcommand{\thetable}{S\@arabic\c@table}
% \renewcommand{\thesection}{Supplemental Note S\@arabic\c@section}
% \makeatother

% \vspace*{1cm}

% {\bf \large Contents}

% % \begin{enumerate}
%     \item \textbf{Supplemental Figures~S1.}
% \end{enumerate}

% \clearpage

% \noindent {\LARGE Supplemental Figures}
